---
title: "SLU Peer Tutoring Program Analysis"
format: revealjs
---



## Data Introduction

This project is an alaysis of the information compiled from peer tutoring request forms from St. Lawrence University.

The data includes the following variables:

-   Duration to fill form (in seconds)	
-   Location
-   Class	Professor	
-   Subject
-   Appointment Start & End Time


## Data Introduction
Created for ethical reasons:

-   Tutor Id	
-   Student id	
-   Class_Year		

```{r}
#| include: false
library(tidyverse)
library(readxl)
library(lubridate)

tutoring = read_excel("data/MAIN_DATA.xlsx")


tutoring = tutoring |>
  mutate(
    start_time = parse_date_time(`Appointment Start Time`, orders = "I:M p"),
    end_time = parse_date_time(`Appointment End Time`, orders= "I:M p"),
    appointment_minutes = as.numeric(end_time - start_time, units = "mins"),
    start_hour = hour(start_time)
  )

tutoring = tutoring |>
  mutate(
    course_combined = if_else(is.na(COURSE), Class, COURSE)) |>
  mutate(
    course_group = case_when(
      is.na(course_combined)~"Missing",
      str_detect(course_combined, "STAT|Stat|stat") ~ "Statistics",
      str_detect(course_combined, "CHEM|Chem|chem") ~ "Chemistry",
      str_detect(course_combined, "BIO|Bio|bio") ~ "Biology",
      str_detect(course_combined,"ECON|Econ|econ|Financial Accounting|ACC 203 Financial Accounting") ~ "Economics",
      str_detect(course_combined,"PSYC|Psyc|psyc") ~ "Psychology",
      str_detect(course_combined,"PHYS|Phys|phys") ~ "Physics",
      str_detect(course_combined,"MATH|Math|math|Calculus 1|Calculus 2|Multivariable Calculus")~ "Mathematics",
      str_detect(course_combined,"CS|Computer|cs") ~ "Computer Science",
      str_detect(course_combined,"SPAN|Spanish|spanish")  ~ "Spanish",
      str_detect(course_combined,"SSES|Environment|General Ecology")  ~ "Environmental Sciences",
      str_detect(course_combined,"PHIL|philosophy")  ~ "Philosophy",
      str_detect(course_combined,"GER|German") ~ "German",
      str_detect(course_combined,"FREN|French|FR") ~ "French",
      str_detect(course_combined,"GOVT")  ~ "Government",
      TRUE~"Other"
    )
  )

top_tutors = tutoring |>
  group_by(`Tutor Id`)|>
  summarise(total_appointments = n()) |>
  arrange(desc(total_appointments)) |>
  slice(1:10)
```


## Question(s) of Interest

 **TUTOR POV** 

**STUDENTS POV**

How is tutoring demand structured in SLU's Peer Tutoring Program, and what patterns emerge in both tutor workload and student engagement?

## Finding 1 -> Tutor Workload Is Concentrated
```{r}
#| echo: false
#cumsum function found in chapter 8 - refine your plot s in our data viz book 
tutor_workload = tutoring |>
  group_by(`Tutor Id`) |>
  summarise(total_appointments = n())|>
  arrange(desc(total_appointments))

tutor_workload = tutor_workload |>
  mutate(
    cum = cumsum(total_appointments)/sum(total_appointments),
    tutor_rank = row_number()
  )

cutoff50 = tutor_workload|>
  filter(cum >=0.5) |>
  slice(1)

ggplot(tutor_workload,
       aes(x=tutor_rank, y = cum)) +
  geom_line()+
  geom_hline(yintercept = 0.5, linetype = "dashed")+
  geom_vline(xintercept = cutoff50$tutor_rank, linetype = "dashed")+
  geom_point(data = cutoff50,
             aes(x = tutor_rank, y =cum),
             size = 3, color = "red")+
  labs(
    title = "Cumulative Share of Tutoring Workload",
    x = "Tutor Rank",
    y = "Cumulative Share of Appointments")+
  theme_minimal(base_size = 24)
```

## Finding 2 -> Top Tutors Specialize by Subject
```{r}
#| echo: false

# Remove "Missing- 1101" which are students that didnt add anything 
tutor_specialization = tutoring |>
  filter(course_group != "Missing",
         course_group != "Other",
         `Tutor Id` %in% top_tutors$`Tutor Id`)|>
  group_by(`Tutor Id`, course_group) |>
  summarise(total = n(), .groups = "drop") |>
  ungroup()

tutor_order = top_tutors|>
  pull(`Tutor Id`)

course_order = tutor_specialization|>
  group_by(course_group)|>
  summarise(total_subject = sum(total))|>
  arrange(desc(total_subject))|>
  pull(course_group)

tutor_specialization = tutor_specialization |>
  mutate(
    `Tutor Id` = factor(`Tutor Id`, levels = tutor_order),
    course_group=factor(course_group, levels = course_order)
  )

# tutor specialization
ggplot(tutor_specialization,
       aes(x=`Tutor Id`, y= course_group, fill = total))+
  geom_tile( color = "white", linewidth = 0.4)+ 
  scale_fill_viridis_c()+
  labs(
    title = "Tutor Specialization Across Subject Areas",
    x = "Tutor Id",
    y = "Course Group",
    fill = "Appointments"
  )+
  theme_minimal(base_size = 26)
```


## Finding 3 -> Students Return to Same Tutor
```{r}
#| echo: false
## returning students per tutor -> how many students returned to the same tutor more than once

same_tutor = tutoring |>
  group_by(`Student id`, `Tutor Id`) |>
  summarise(visits_tutor = n()) |>
  filter(visits_tutor > 1) |>
  group_by(`Tutor Id`) |>
  summarise(
    same_return = n())|>
  arrange(desc(same_return))



# Observing it with rate

student_tutors = tutoring |>
  group_by(`Tutor Id`,`Student id`) |>
  summarise()|>
  group_by(`Tutor Id`) |>
  summarise(total_students =n())

tutors_rate = same_tutor |>
  left_join(student_tutors, by ="Tutor Id") |>
  mutate(return_rate = same_return/total_students)|>
  arrange(desc(return_rate))

tutors_rate |>
  filter(`Tutor Id` %in% top_tutors$`Tutor Id`)|>
  ggplot(aes(x = reorder(`Tutor Id`, return_rate),
             y = return_rate)) +
  geom_col(fill = "brown4")+
  coord_flip() +
  labs(
    title = "Rate of Students Returning to their Assigned Tutor",
    x = "Tutor Id",
    y = "Return Rate")+
  theme_minimal(base_size = 24)
```


## Finding 4 -> Repeat Support Happens in Specific Subjects
```{r}
#| echo: false
# returning behavior: same course same tutor

stu_return = tutoring |>
  group_by(`Student id`, course_group, `Tutor Id`)|>
  summarise(visit_sameCT = n())|>
  filter(visit_sameCT>1)

stu_return |>
  group_by(course_group)|>
  summarise(total_sameCT = n())|>
  filter(course_group != "Missing")|>
  arrange(desc(total_sameCT))|>
  ggplot(aes(x = reorder(course_group, total_sameCT),
             y = total_sameCT,
             fill = total_sameCT))+
  geom_col()+
  coord_flip()+
  scale_fill_viridis_c() +
  labs(
    title = "Repeated Same Course and Same Tutor Returns by Subject",
    x = "Subject Area",
    y = "Number of Repeat Student-Tutor Pairings",
    fill = "Returns"
  )+
  theme_minimal(base_size = 24)
```

## Finding 5 -> Tutor Demands Varies by Class Year
```{r}
#| echo: false
# class year usage

tutoring = tutoring |>
  mutate(Class_Year_Code =as.numeric(Class_Year_Code))

tutoring = tutoring |>
  filter(!is.na(Class_Year_Code))

tutoring = tutoring |>
  mutate( grad_year = Class_Year_Code + 4)

class_year = tutoring |>
  group_by(grad_year)|>
  summarise(total_requests = n())|>
  filter(total_requests >10)|>
  arrange(grad_year)

ggplot(class_year, aes(x=as.factor(grad_year), y=total_requests))+
  geom_segment(aes(x = as.factor(grad_year), xend = as.factor(grad_year), y = 0, yend = total_requests), color = "#8B5A2B")+
  geom_point(size = 4, color = "#8B1E1E")+
  labs(
    title = "Tutoring Requests by Student Class Year",
    x = "Class Year Code",
    y = "Number of Requests")+
  theme_minimal(base_size = 24)+
  coord_flip()
```


## What Now?
-   Balance Tutor Workload
-   Strengthen subject-based matching
-   Support repeat tutoring relationships
-   Improve data collection