---
title: "Tutors"
format: html
---

# Frequent course tutor gets requested
# return students and what courses 
# tutor with more return students 
# evaluations
# location spot 
# appointments times usual ones
# meeting duration -> hovering??? i want to know if there are some subjects that take longer 


```{r}
library(tidyverse)
library(readxl)
library(lubridate)
```

```{r}
tutoring = read_excel("data/MAIN_DATA.xlsx")
head(tutoring)
```

```{r}
tutoring = tutoring |>
  mutate(
    start_time = parse_date_time(`Appointment Start Time`, orders = "I:M p"),
    end_time = parse_date_time(`Appointment End Time`, orders= "I:M p"),
    appointment_minutes = as.numeric(end_time - start_time, units = "mins"),
    start_hour = hour(start_time)
  )
```


```{r}
# tutor workload
top_tutors = tutoring |>
  group_by(`Tutor Id`)|>
  summarise(total_appointments = n()) |>
  arrange(desc(total_appointments)) |>
  slice(1:10)

ggplot(top_tutors,
       aes(x = reorder(`Tutor Id`, total_appointments),
           y = total_appointments)) +
  geom_col()+
  coord_flip() +
  labs(
    title = "Top Tutors by Number of Appointments",
    x = "Tutor ID",
    y = "Number of Appointments"
  )

tutoring |>
  group_by(`Tutor Id`) |>
  summarise(total_appointments = n()) |>
  ggplot(aes(x = total_appointments)) +
  geom_histogram(binwidth = 5) +
  labs(
    title = "Distribution of Appointments Across Tutors",
    x = "Number of Appointments",
    y = "Number of Tutors"
  ) +
  theme_minimal()
```




```{r}

tutoring = tutoring |>
  mutate(
    course_combined = if_else(is.na(COURSE), Class, COURSE)) |>
  mutate(
    course_group = case_when(
      is.na(course_combined)~"Missing",
      str_detect(course_combined, "STAT|Stat|stat") ~ "Statistics",
      str_detect(course_combined, "CHEM|Chem|chem") ~ "Chemistry",
      str_detect(course_combined, "BIO|Bio|bio") ~ "Biology",
      str_detect(course_combined,"ECON|Econ|econ|Financial Accounting|ACC 203 Financial Accounting") ~ "Economics",
      str_detect(course_combined,"PSYC|Psyc|psyc") ~ "Psychology",
      str_detect(course_combined,"PHYS|Phys|phys") ~ "Physics",
      str_detect(course_combined,"MATH|Math|math|Calculus 1|Calculus 2|Multivariable Calculus")~ "Mathematics",
      str_detect(course_combined,"CS|Computer|cs") ~ "Computer Science",
      str_detect(course_combined,"SPAN|Spanish|spanish")  ~ "Spanish",
      str_detect(course_combined,"SSES|Environment|General Ecology")  ~ "Environmental Sciences",
      str_detect(course_combined,"PHIL|philosophy")  ~ "Philosophy",
      str_detect(course_combined,"GER|German") ~ "German",
      str_detect(course_combined,"FREN|French|FR") ~ "French",
      str_detect(course_combined,"GOVT")  ~ "Government",
      TRUE~"Other"
    )
  )

tutoring |>
  filter(course_group == "Other") |>
  group_by(if_else(is.na(COURSE), Class, COURSE))|>
  summarise(total = n())|>
  arrange(desc(total))|>
  slice(1:30)

# Remove "Missing- 1101" which are students that didnt add anything 
tutor_specialization = tutoring |>
  filter(course_group != "Missing",
         course_group != "Other",
         `Tutor Id` %in% top_tutors$`Tutor Id`)|>
  group_by(`Tutor Id`, course_group) |>
  summarise(total = n(), .groups = "drop") |>
  ungroup()

tutor_order = top_tutors|>
  pull(`Tutor Id`)

course_order = tutor_specialization|>
  group_by(course_group)|>
  summarise(total_subject = sum(total))|>
  arrange(desc(total_subject))|>
  pull(course_group)

tutor_specialization = tutor_specialization |>
  mutate(
    `Tutor Id` = factor(`Tutor Id`, levels = tutor_order),
    course_group=factor(course_group, levels = course_order)
  )
```

```{r}
# tutor specialization
ggplot(tutor_specialization,
       aes(x=`Tutor Id`, y= course_group, fill = total))+
  geom_tile( color = "white", linewidth = 0.4)+ 
  scale_fill_viridis_c()+
  labs(
    title = "Tutor Specialization Across Subject Areas",
    x = "Tutor Id",
    y = "Course Group",
    fill = "Appointments"
  )+
  theme_minimal()
```

## appointment times
```{r}
ggplot(tutoring, aes(x = start_hour))+
  geom_histogram(binwidth = 1)+
  labs(
    title = "Distribution of Appointment Start Times",
    x = "Hour of the Day",
    y = "Number of Appointments"
  )+
  theme_minimal()
```

## session duration
```{r}
tutoring |>
  filter(`Tutor Id` %in% top_tutors$`Tutor Id`,
         !is.na(start_hour))|>
  ggplot(aes(x = start_hour))+
  geom_histogram(binwidth = 1)+
  facet_wrap(~`Tutor Id`)+
  labs(
    title = "Appointment Start Times for Top Tutors",
    x = "Hour of the Day",
    y= "Number of Appointments"
  ) +
  theme_minimal()


### important to bring up this is a disadvantage and probably something to reinforce
tutoring |> 
  filter(`Tutor Id` %in% top_tutors$`Tutor Id`)|>
  group_by(`Tutor Id`)|>
  summarise(
    total_rows = n(),
    non_missing_times = sum(!is.na(start_hour))
  )
```


## returning students per tutor -> how many students returned to the same tutor more than once
```{r}
same_tutor = tutoring |>
  group_by(`Student id`, `Tutor Id`) |>
  summarise(visits_tutor = n()) |>
  filter(visits_tutor > 1) |>
  group_by(`Tutor Id`) |>
  summarise(
    same_return = n())|>
  arrange(desc(same_return))

same_tutor

same_tutor|>
  filter(`Tutor Id` %in% top_tutors$`Tutor Id`)|>
  ggplot(aes(x = reorder(`Tutor Id`, same_return),
             y = same_return))+
  geom_col()+
  coord_flip()+
  labs(title = "Students Who Returned to the Same Tutor",
       x = "Tutor Id",
       y = "Number of Students")+
  theme_minimal()
```

# Observing it with rate
```{r}
student_tutors = tutoring |>
  group_by(`Tutor Id`,`Student id`) |>
  summarise()|>
  group_by(`Tutor Id`) |>
  summarise(total_students =n())

tutors_rate = same_tutor |>
  left_join(student_tutors, by ="Tutor Id") |>
  mutate(return_rate = same_return/total_students)|>
  arrange(desc(return_rate))

tutors_rate |>
  filter(`Tutor Id` %in% top_tutors$`Tutor Id`)|>
  ggplot(aes(x = reorder(`Tutor Id`, return_rate),
             y = return_rate)) +
  geom_col()+
  coord_flip() +
  labs(
    title = "Rate of Students Returning to their Assigned Tutor",
    x = "Tutor Id",
    y = "Return Rate")+
  theme_minimal()
```

# worload concentration -> is work concentrated only in few tutors

```{r}
#cumsum function found in chapter 8 - refine your plot s in our data viz book 
tutor_workload = tutoring |>
  group_by(`Tutor Id`) |>
  summarise(total_appointments = n())|>
  arrange(desc(total_appointments))

tutor_workload = tutor_workload |>
  mutate(
    cum = cumsum(total_appointments)/sum(total_appointments),
    tutor_rank = row_number()
  )

cutoff50 = tutor_workload|>
  filter(cum >+0.5) |>
  slice(1)

ggplot(tutor_workload,
       aes(x=tutor_rank, y = cum)) +
  geom_line()+
  geom_hline(yintercept = 0.5, linetype = "dashed")+
  geom_vline(xintercept = cutoff50$tutor_rank, linetype = "dashed")+
  geom_point(data = cutoff50,
             aes(x = tutor_rank, y =cum),
             size = 3)+
  labs(
    title = "Cumulative Share of Tutoring Workload",
    x = "Tutor Rank",
    y = "Cumulative Share of Appointments")+
  theme_minimal()
```


## are top tutors responsible for most of the deman in certain subjects? ---> i dont think this one is ccurate

```{r}
subject_tt = tutoring |>
  filter(`Tutor Id` %in% top_tutors$`Tutor Id`,
         course_group!="Missing",
         course_group!="Other")|>
  group_by(course_group)|>
  summarise(
    top_tutors_appt = n()
  )

subject_all = tutoring |>
  filter(course_group!="Missing",
  course_group!="Other")|>
  group_by(course_group)|>
  summarise(
    total_appointments= n()
  )

subject_share = subject_tt |>
  left_join(subject_all, by = "course_group")|>
  mutate(
    share_top = top_tutors_appt/total_appointments
  )

ggplot(subject_share, aes(x = reorder(course_group, share_top),
                          y= share_top)) +
  geom_col()+
  coord_flip()+
  labs(title = "Share of Appointments Handled by Top Tutors",
       x = "Subject Area",
       y = "Proportion")+
  theme_minimal()
```


# do tutors with more appointments are they the ones that have the largest returning number of students? ---> issues with same_tutor_rate 
```{r}
tutor_combined = tutor_workload |>
  left_join(same_tutor, by = "Tutor Id")

ggplot(tutor_combined, aes(x = total_appointments, y = return_rate))+
  geom_point()+
  labs(
    title = "Tutor Workload vs Same Tutor Return Rate",
    x = "Total Appointments",
    y= "Return Rate")+
  theme_minimal()

## attempt with not using rate?
  
```

# students section 

```{r}
# class year usage

tutoring = tutoring |>
  mutate(Class_Year_Code =as.numeric(Class_Year_Code))

tutoring = tutoring |>
  filter(!is.na(Class_Year_Code))

tutoring = tutoring |>
  mutate( grad_year = Class_Year_Code + 4)

class_year = tutoring |>
  group_by(grad_year)|>
  summarise(total_requests = n())|>
  filter(total_requests >10)|>
  arrange(grad_year)

ggplot(class_year, aes(x=as.factor(grad_year), y=total_requests))+
  geom_segment(aes(x = as.factor(grad_year), xend = as.factor(grad_year), y = 0, yend = total_requests), color = "blue")+
  geom_point(size = 4, color = "green")+
  labs(
    title = "Tutoring Requests by Student Class Year",
    x = "Class Year Code",
    y = "Number of Requests")+
  theme_minimal()
```


```{r}
# Duration to fill out the form - exposing how easy of a resource it is and can we improve

tutor_sec = tutoring |>
  filter(`Duration to fill form (in seconds)`<=2000)

ggplot(tutoring, aes(x = as.factor(grad_year),
                     y = `Duration to fill form (in seconds)`))+
  geom_boxplot(fill = "red", alpha = 0.7) +
  coord_cartesian(ylim = c(0, 1000))+
  labs(
    title = " Time Spent Filling Out the Tutoring Request Form",
    x = "Class Year",
    y = "Duration to Fill Form in Seconds")+
  theme_minimal()
  
```


```{r}
# returning behavior: same course same tutor

stu_return = tutoring |>
  group_by(`Student id`, course_group, `Tutor Id`)|>
  summarise(visit_sameCT = n())|>
  filter(visit_sameCT>1)

stu_return |>
  group_by(course_group)|>
  summarise(total_sameCT = n())|>
  arrange(desc(total_sameCT))|>
  ggplot(aes(x = reorder(course_group, total_sameCT),
             y = total_sameCT,
             fill = total_sameCT))+
  geom_col()+
  coord_flip()+
  scale_fill_viridis_c() +
  labs(
    title = "Repeated Same Course and Same Tutor Returns by Subject",
    x = "Subject Area",
    y = "Number of Repeat Student-Tutor Pairings",
    fill = "Returns"
  )+
  theme_minimal()
```

```{r}
ggplot(tutoring,
       aes(x = LocationLongitude,
           y = LocationLatitude,
           color = course_group)) +
  geom_point(alpha = 0.4) +
  scale_color_viridis_d()+
  labs(
    title = "Location Patterns for Tutoring",
    x = "Longitude",
    y = "Latitude",
    color = "Subject Area"
  )+
  theme_minimal()
```

to get some ideas of color i could use i input on gemini "give me list of color blind friendly colors for scarlett and brown"

```{r}
return_data = tutoring |>
  group_by(`Student id`, `Tutor Id`)|>
  summarise(visits = n())|>
  mutate(return_same_t = if_else(visits > 1, 1,0))

model_data = tutoring |>
  left_join(return_data, by = c("Student id", "Tutor Id"))

model_data = model_data|>
  filter(!is.na(start_hour),
         !is.na(course_group))

model = glm(return_same_t~ start_hour + course_group,
            data = model_data,
            family = "binomial")

summary(model)
```

